\Large {\bf Protocol-Guided Analysis of Post-silicon Traces Under Limited Observability}

### 2015-12-21 paper.tex
added one more reference

%% \author{\large Hao Zheng$^1$, Yuting Cao$^1$, Sandip Ray$^2$, Jin Yang$^2$ \\
%% $^1$Dept. of Computer Science and Eng., University of South Florida, Tampa, Fl 33620. USA. \\
%% $^2$Strategic CAD Labs, Intel Corporation, Hillsboro, OR 97124.  USA.}

@@

Post-silicon validation makes use of pre-production silicon
integrated circuit (IC) to ensure that the fabricated system
works as desired under actual operating conditions with real
software.  Since the silicon executes at target clock speed,
post-silicon executions are billions of times faster than
RTL simulations, and even provide speed-up of several orders
of magnitude over other pre-silicon platforms (\eg, FPGA,
system-level emulation, etc.).  This makes it possible to
explore deep design states which cannot be exercised in
pre-silicon, and identify errors missed during pre-silicon
validation and debug.  Post-silicon validation is a critical
component of the design validation life-cycle for modern
microprocessors and SoC designs.  Unfortunately, it is also
a highly complex component, performed under aggressive
schedules and accounting for more than $50\%$ of the overall
design validation cost.  Consequently, it is crucial to
develop techniques for streamlining and automating
post-silicon validation activities.

@@

\begin{figure*}[tb]
\begin{center}
\begin{tabular}{cc}
\begin{minipage}{3in}
\includegraphics[height=2in]{figures/bpmn-flow-ex}
\end{minipage}
&
\begin{minipage}{3in}
\resizebox{2.8in}{!}{
\begin{tikzpicture}[node distance=3cm, auto,>=latex', thick]
\tikzset{rectbox/.style={rectangle, draw, align=flush center,thick, minimum size = 6mm}}
\tikzset{circ/.style={circle, draw,thick}}
\tikzset{terminal/.style={circle, draw,line width=.8mm}}
\tikzset{weight3/.style={line width=.3mm}}
	% Nodes:
	\node[circ]		(p1) 		at (0,0) {$p_1$};
	\node[rectbox]	(t1)	at	($(p1) + (0,-1.1)$) {$t_1:\langle {\tt Driver:Device:Load} \_fw\rangle$};
	\node[circ] 	(p2) 		at ($(t1) + (0,-1.1)$) {$p_2$};
	\node[rectbox]	(t2)	at	($(p2) + (0,-1.1)$) {$t_2:\langle {\tt Device:CE:Auth\_req} \rangle$};
	\node[circ] 	(p3) 		at ($(t2) + (0,-1.1)$) {$p_3$};
	\node[rectbox]	(t3)	at	($(p3) + (0,-1.1)$) {$t_3:\langle {\tt CE:Device:Auth\_resp}\rangle$};
	\node[circ] 	(p4) 		at ($(t3) + (0,-1.1)$) {$p_4$};
	\node[rectbox]	(t4)	at	($(p4) + (-3,-1.1)$) {$t_4:\langle {\tt Device:Driver:Report} \rangle$};
	\node[rectbox]	(t5)	at	($(p4) + (3,-1.1)$) {$t_5:\langle {\tt Device:CE:Ack} \rangle$};
	\node[circ,fill=black] 	(term1) 		at ($(t4) + (0,-1.1)$) {\textcolor{white}{$p_5$}};
	\node[circ,fill=black] 	(term2) 		at ($(t5) + (0,-1.1)$) {\textcolor{white}{$p_6$}};

@@

An SoC design involves integration of a number of IPs that
communicate through complex protocols.  Such system-level
protocols are typically specified in architecture documents
as message flow diagrams.  For this paper, we use the words
``protocol'' and ``flow'' interchangeably.
Fig~\ref{flow-spec-ex}(a) shows one diagram for a protocol
to authenticate and load a firmware during system boot for
firmware upgrade.  During validation, the system under debug
(SUD) exercises some complex system-level use-case which
involves interleaved execution of possibly a large number of
such flows.  A trace of a small number of hardware signals
is then shipped off-chip analysis.  The off-chip analysis
includes two broad phases: (1)~trace abstraction, and
(2)~trace interpretation.  Trace abstraction maps the
hardware trace into higher-level architectural constructs,
\eg, messages, operations, etc.: a message such as {\tt
  Authorization request} may be implemented in hardware
through a Boolean or temporal combination of specific
hardware signals in the NoC fabric between {\tt Device} and
{\tt CE}, \eg, as a sequence containing a header, a specific
value of a sequence of data words, etc.  We will refer to
such architectural constructs as {\em protocol events} or
{\em flow events}.  Note that due to limited observability,
it may not be possible to map a given set of (observed)
hardware signals uniquely to a flow event.  Finally, the
trace may a result from several instances of the same
protocol executing concurrently, \eg, a firmware
authentication protocol may be invoked when another instance
of the protocol has not completed.

@@

\noindent 
{\bf Notations and formalization.}  We assume that the set
of system flows is provided as a collection ${\vec{F}}$ of
LPNs.  We define a {\em flow execution scenario} to be a
set $\{(F_{i,j}, s_{i,j})\}$ where $F_{i,j}$ is the
$j$th instance of flow $F_i \in {\vec{F}}$, and $s_{i,j}$ is
a state of $F_{i,j}$.  A flow execution scenario indicates
the set of protocols and the number of instances of a
particular protocol are activated and their corresponding
current states.  Since we assume full observability, we view
an {\em observed trace} $\rho = e_1e_2\ldots e_n$ is a
sequence of events.  Given an observed trace $\rho$, the
goal of of trace interpretation is to construct a set of
candidate flow execution scenarios whose execution can
create the sequence of events in $\rho$.  We call such
execution scenarios {\em compliant} with $\rho$.  Let
$\mathit{accept(F_{i,j}, s_{i,j}, e)}$ be a function that
determines if event $e$ can be emitted by $F_{i,j}$ in state
$s_{i,j}$.  Formally, $\mathit{accept(F_{i,j}, s_{i,j}, e)}$
returns $(F_{i,j}, s'_{i,j})$ where $s'_{i,j} = (s_{i,j} - \bullet t)
\cup t\bullet$ if there exists a transition
$t$ in $F_i$ such that $L(t) = e$ and $\bullet t \subseteq
s_{i,j}$.  It returns $\emptyset$ if no such $t$
exists in $F_i$.

Given an obsered trace $\rho$ and the set $\vec{F}$ of LPNs,
Algorithm~\ref{algo:compliance} provides a basic procedure for
computing a set of compliant flow execution scenarios.
\begin{algorithm}[ht]
\DontPrintSemicolon
Create an empty scenario $scen$\;
$\mathit{Scen} = \{scen\}$\;
\ForEach {$h, \; 1 \leq h \leq n $} {
	$\mathit{found} \gets {\tt true}$ \;
	$Scen' = \emptyset$\;
	\ForEach {$scen \in Scen$} {
  		\ForEach {$(F_{i,j}, s_{i,j}) \in \mathit{scen_1}$} {
    		\If {$\mathit{accept}(F_{i,j}, s_{i,j}, e_h) = (F_{i,j}, s'_{i,j})$} {
				Let $scen'$ be a copy of $scen$\;
      		$\mathit{scen'} \gets (\mathit{scen'} - (F_{i,j}, s_{i,j})) \cup (F_{i,j}, s'_{i,j})$\;
      		$\mathit{Scen'} \gets \mathit{scen'} \cup \mathit{Scen'}$\;
      		$\mathit{found} \gets {\tt false}$ \;
    		}
  		}
  		\ForEach {$F_i \in \vec{F}$} {
      	create a new instance $F_{i, j+1}$ \;
      	\If {$\mathit{accept}(F_{i,j+1},\mathit{init}_{i,j+1}, e_h) = (F_{i,j+1}, s'_{i,j+1})$} {
    			Let $scen'$ be a copy of $scen$\;
				$\mathit{scen'} \gets \mathit{scen'} \cup (F_{i,j+1}, s'_{i,j+1})$ \;
				$\mathit{Scen'} \gets \mathit{scen'} \cup \mathit{Scen'}$\;
        		$\mathit{found} \gets {\tt false}$ \;
      	}
    	}
	}
  \If {$\mathit{found} == {\tt true}$} {
    \Return ${\tt Inconsistent}$\;
  }
  $Scen = Scen'$\;
}
\Return $\mathit{Scen}$ \;
\caption{$\textsc{Check-Compliance}(\vec{F}, \, \rho)$}
\label{algo:compliance}
\end{algorithm}
The algorithm operates by keeping track (in variable
$\mathit{Scen}$) of a set of candidate flow execution scenarios
compliant with each prefix of $\rho$.  At each iteration,
for each event $e_h$ in the observed trace, we
update $Scen$ by either extending a member of
$\mathit{scen}$ or initiating a new protocol instance for 
each $scen \in Scen$ with respect to $e_h$ in every possible way.
If $e_h$ cannot be emitted by any existing or new flow instances, 
then we report that the trace is {\tt  inconsistent}, \ie, 
there is no possible interleaving of
the protocol instances from $\vec{F}$ that is compliant with
$\rho$.

To illustrate the basic idea, consider the system flow in
Fig.~\ref{flow-spec-ex}(b), which we will call $F_1$.
Suppose that the following flow trace is abstracted from an
observed signal trace of the system.
\[
	t_1\;t_2\;t_1\;t_2\;t_3\;t_3\;t_4\;t_5\;t_5\;t_4\ldots
\]  
Here transition names in the LPN are used to represent the 
flow events in the trace.  The first four events results in
the following flow execution scenario
\[
	\{(F_{1,1}, \{p_3\}),~(F_{1,2}, \{p_3\})\}.
\]
For the first event $t_3$, it results in two execution scenarios 
below depending on which flow instance emits $t_3$.
\[
\begin{array}{l}
	\{(F_{1,1}, \{p_4\}),~(F_{1,2}, \{p_3\})\} \\
	\{(F_{1,1}, \{p_3\}),~(F_{1,2}, \{p_4\})\}.
\end{array}
\]
After handing the next event $t_3$, the above two execution scenarios
are reduced to the one as shown below.
\[
	\{(F_{1,1}, \{p_4\}),~(F_{1,2}, \{p_4\})\}.
\]
Using Algorithm~\ref{algo:compliance} to handle the
remaining four events, the following execution scenario is
derived.
\[
	\{(F_{1,1}, \{p_5, p_6\}),~(F_{1,2}, \{p_5, p_6\})\}
\]
%% Given a trace of flow events $\rho = e_1e_2\ldots e_n$, the
%% trace interpretation algorithm starts with an empty set of
%% of flow execution scenario $\mathit{Scen} = \emptyset$.
%% Then, for each $e_h$ where $1 \leq h \leq n$ starting $h=1$,
%% and for each $\mathit{scen} \in \mathit{Scen}$, the
%% following two steps are performed.
%% \begin{description}
%% \item[Step 1]~~For each $(F_{i,j}, s_{i,j}) \in
%%   \mathit{scen}$, if $\mathit{accept}(F_{i,j}, s_{i,j}, e_h)
%%   = (F_{i,j}, s'_{i,j})$, create a new scenario
%%   $\mathit{scen}' = (\mathit{scen} - (F_{i,j}, s_{i,j}))
%%   \cup (F_{i,j}, s'_{i,j})$, which is added into
%%   $\mathit{Scen}'$.

@@

In general, a signal trace of partial observability
corresponding a set of traces of flow events due to the
ambiguous interpretation of signal events.  In the
following, we discuss two cases for trace abstraction on
partial observability: mapping a single signal event to a
flow event or mapping a sequence of signal events to a flow
event.  A signal event is defined as a state on or an
assignment to a set of signals.

@@

where $\textvisiblespace$ denotes signal events that are not
mapped to any flow events.  Note that the above abstraction
leads to three distinct flow traces as the middle three
correspond to the same trace of flow events.

It is clear from above that a partial trace is viewed as a
set of flow traces, and Algorithm~\ref{algo:compliance} can
be suitably extended to work with flow traces to obtain the
set of candidate flows.  However, applicability of the
algorithm in practice can be gated because the number of
potential flow execution scenarios generated under partial
observability may be enormous.  Note that this is not a
limitation of the algorithm; if the observability of
critical events is poor there simply {\em are} too many flow
execution scenarios compliant with the observed trace.
Nevertheless, we need to address the issue to make trace
interpretation (whether automatic or not) practicable.
There are two potential approaches: (1)~better selection of
post-silicon trace observability, and (2)~use of system
insights during validation.  Trace signal selection itself
is an important and orthogonal topic~\cite{nicolici,basu},
and a detailed discussion of it is out of scope of this
paper.  However, we briefly describe how the debuggers'
insights of a system's architecture can help to address the
complexity issue in the trace trace interpretation.

@@

Post-silicon validation is performed by debuggers with deep
knowledge about the system's architecture and
microarchitecture, and the test environment.  Two key
insights are (1) the maximal number of instances of a flow
activated in the test environment, and (2)~the mutual
relationship between two flows.  For example, the test
environment may not permit multiple instances of firmware
authentication to operate concurrently, or a flow involving
audio and Web browsing to initiate until the flows
participating in boot are completed.  Our framework permits
incorporating such insights as constraints in trace
analysis; flow execution scenarios that violate these
constraints are ignored.  Apart from the explosion problem,
this permits the debugger to quickly filter out
uninteresting combination of flows and focus on interesting
interleavings.

@@

\begin{figure} 
\centerline{
\includegraphics[width=3.3in]{figures/Fig4.png}}
\caption{SoC platform structure.}
\label{SoC}
\end{figure}

In order to find out the efficiency of the trace analysis
method for more realistic examples, in the case study
described in this section, a more detailed transaction level
model of a SoC is constructed within the GEM5
environment~\cite{Binkert2011}.  This SoC model consists of
two ARM Cortex-A9 cores, each of which contains two separate
16KB data and instruction caches.  The caches are connected
to a 1GB memory through a memory bus model.  The
architecture of this SoC model is shown in Fig.~\ref{SoC}.

@@

For this model, we consider the flow specifications
describing the cache coherence protocols supported in GEM5
that is used to build the model in Fig~\ref{SoC}.  These
flow specifications describe data/instruction read
operations and data write operations initiated from CPUs.
Three such flows describe the cache coherent protocols for
each CPU.  As there are two CPUs in this model, there are
actually six such flows considered for this model.  These
flow specifications are derived from the cache coherence
protocols supported in GEM5.\footnote{The GEM5 cache
  coherence protocols can be found at
  \url{http://www.gem5.org/docs/html/gem5MemorySystem.html}.}

@@

We wrote two simple concurrent programs --- one for each CPU
--- to exercise the flows.  They read numbers from a file,
perform some operations on these numbers, and store the
results back to the file.  How GEM5 supports shared memory
multi-threaded program execution is unclear.  Therefore,
there is no data shared in both caches in this test.
Furthermore, GEM5 does not support true concurrency.  When
there are two programs running on the CPUs, GEM5 alternates
the executions between the two CPUs.  To simulate
asynchronous concurrency with the interleaving semantics,
those two simple programs are instrumented with
pseudo-blocking commands, one placed before each statement.
A pseudo blocking command includes a random number generator
that returns either $0$ or $1$ and a loop that only exits
when the returned random number is $0$.

@@

The time taken to remove the irrelevant messages from the
trace is negligible.  The total runtime and the peak memory
usage for the trace analysis algorithm to finish the reduced
trace are 3 seconds and 12MB, respectively.  From the trace,
Table~\ref{table-case-2} shows the number of instances
extracted for the six flows describing cache coherent
read/write operations initiated from both CPUs.
\begin{table}[tb]
\caption{The number of flow instances derived by the trace analysis with the full observability.}
\begin{center}
\begin{tabular}{|l|c|}
\hline
Flows & $\#$Instances \\
\hline
\hline
CPU1 Data Read			&  $17582$\\
CPU1 Instruction Read		&  $4002$\\
CPU1 Write				&  $3370$\\
\hline
CPU2 Data Read			&  $17386$\\
CPU2 Instruction Read		&  $3955$\\
CPU2 Write				&  $3308$\\
\hline
\end{tabular}
\end{center}
\label{table-case-2}
\end{table}%

In the second experiment, partial observability is taken into account
with the four monitors attached to the links between two CPUs and their
caches are disabled. Then, the trace is generated by the
remaining five monitors from the SoC model executing the
same program. After removing all non-observable messages,
the trace only contains 15089 messages. The numbers of the
flow instances extracted by the trace analysis method are
shown in Table~\ref{table-par-obs}.  From these results, the
numbers of the flow instances are dropped significantly
compared to the results extracted from the trace with the
full observability as shown in
Table~\ref{table-case-2}. This difference is due to that
some communications occurred in the system when executing
the program involve the CPUs and their corresponding caches
only, and the traffic on the links between the CPUs and
their corresponding caches is not observable. Therefore, the
instances of the flow specifications characterizing these
communications do not exist in the trace. In other words,
all extracted flow instances in Table~\ref{table-par-obs}
characterize the communications that pass through the memory
bus in the system model.  The runtime and memory usage as shown in
the third column in Table~\ref{table-results} are
similar to those for analyzing the trace of the full
observability.

@@

In the third experiment, further partial observability is taken into consideration.  In this experiment, only the five links involving the memory bus are still considered.  However, an assumption is made that all events passing the same link are not distinguishable due to the limitation of the observability.  More specifically, the monitors are modified such that whenever an event is captured on one of the links, it dumps a set of events passing through the same link into the trace file.  Therefore, each line of the trace file corresponds to a set of events.  After applying the trace analysis to this trace,  a total of 13944 flow execution scenarios are extracted.    This large number, compared to the number of extracted execution scenarios shown in the Table~\ref{table-case-2} and \ref{table-par-obs}, is due to the ambiguous interpretation of the events with limited observability.

The whole experiment takes about $15$ minutes and $420$~MB to finish as shown in column~4
in Table~\ref{table-results}, significantly higher than the numbers for analyzing traces where there is no ambiguity in the observed events.  This is due to the fact that a trace of ambiguous events is in fact a set of traces of original events, which lead to large numbers of execution scenarios either during or at the end of the analysis.  In this experiment, the peak number of execution scenarios during the analysis process is $70384$.   Compared to the final number, many of the intermediate execution scenarios are invalid, and removed eventually.  However, controlling the number of intermediate execution scenarios during the trace analysis is critical in order for the analysis to be tractable.  Here, insights from validators can help.

>>>>

 \author{\large Hao Zheng$^1$, Yuting Cao$^1$, Sandip Ray$^2$, Jin Yang$^2$ \\
 $^1$Dept. of Computer Science and Eng., University of South Florida, Tampa, Fl 33620. USA. \\
 $^2$Strategic CAD Labs, Intel Corporation, Hillsboro, OR 97124.  USA.}

@@

Post-silicon validation makes use of pre-production silicon
integrated circuit (IC) to ensure that the fabricated system
works as desired under actual operating conditions with real
software.  Since the silicon executes at target clock speed,
post-silicon executions are billions of times faster than
RTL simulations, and even provide speed-up of several orders
of magnitude over other pre-silicon platforms (\eg, FPGA,
system-level emulation, etc.).  This makes it possible to
explore deep design states which cannot be exercised in
pre-silicon, and identify errors missed during pre-silicon
validation and debug.  Post-silicon validation is a critical
component of the design validation life-cycle for modern
microprocessors and SoC designs.  Unfortunately, it is also
a highly complex component, performed under aggressive
schedules and accounting for more than $50\%$ of the overall
design validation cost~\cite{Patra2007}.  Consequently, it is crucial to
develop techniques for streamlining and automating
post-silicon validation activities.

@@

\begin{figure*}[tb]
\begin{center}
\begin{tabular}{cc}
\begin{minipage}{3in}
\includegraphics[height=2in]{figures/bpmn-flow-ex}
\end{minipage}
&
\begin{minipage}{3in}
\resizebox{2.4in}{!}{
\begin{tikzpicture}[node distance=3cm, auto,>=latex', thick]
\tikzset{rectbox/.style={rectangle, draw, align=flush center,thick, minimum size = 6mm}}
\tikzset{circ/.style={circle, draw,thick}}
\tikzset{terminal/.style={circle, draw,line width=.8mm}}
\tikzset{weight3/.style={line width=.3mm}}
	% Nodes:
	\node[circ]		(p1) 		at (0,0) {$p_1$};
	\node[rectbox]	(t1)	at	($(p1) + (0,-1.1)$) {$t_1:\langle {\tt Driver:Device:Load} \_fw\rangle$};
	\node[circ] 	(p2) 		at ($(t1) + (0,-1.1)$) {$p_2$};
	\node[rectbox]	(t2)	at	($(p2) + (0,-1.1)$) {$t_2:\langle {\tt Device:CE:Auth\_req} \rangle$};
	\node[circ] 	(p3) 		at ($(t2) + (0,-1.1)$) {$p_3$};
	\node[rectbox]	(t3)	at	($(p3) + (0,-1.1)$) {$t_3:\langle {\tt CE:Device:Auth\_resp}\rangle$};
	\node[circ] 	(p4) 		at ($(t3) + (0,-1.1)$) {$p_4$};
	\node[rectbox]	(t4)	at	($(p4) + (-3,-1.1)$) {$t_4:\langle {\tt Device:Driver:Report} \rangle$};
	\node[rectbox]	(t5)	at	($(p4) + (3,-1.1)$) {$t_5:\langle {\tt Device:CE:Ack} \rangle$};
	\node[circ,fill=black] 	(term1) 		at ($(t4) + (0,-1.1)$) {\textcolor{white}{$p_5$}};
	\node[circ,fill=black] 	(term2) 		at ($(t5) + (0,-1.1)$) {\textcolor{white}{$p_6$}};

@@

An SoC design involves integration of a number of IPs that
communicate through complex protocols.  Such system-level
protocols are typically specified in architecture documents
as message flow diagrams.  For this paper, we use the words
``protocol'' and ``flow'' interchangeably.
Fig~\ref{flow-spec-ex}(a) shows one diagram for a protocol
to authenticate and load a firmware during system boot for
firmware upgrade.  During validation, the system under debug
(SUD) exercises some complex system-level use-case which
involves interleaved execution of possibly a large number of
such flows.  A trace of a small number of hardware signals
is then shipped off-chip analysis.  The off-chip analysis
includes two broad phases: (1)~trace abstraction, and
(2)~trace interpretation.  Trace abstraction maps the
hardware trace into higher-level architectural constructs,
\eg, messages, operations, etc.: a message such as {\tt
  Authorization request} may be implemented in hardware
through a Boolean or temporal combination of specific
hardware signals in the NoC fabric between {\tt Device} and
{\tt CE}, \eg, as a sequence containing a header, a specific
value of a sequence of data words, etc.  We will refer to
such architectural constructs as {\em protocol events} or
{\em flow events}.  Note that due to limited observability,
it may not be possible to map a given set of (observed)
hardware signals uniquely to a flow event.  Finally, the
trace may be a result from several instances of the same
protocol executing concurrently, \eg, a firmware
authentication protocol may be invoked when another instance
of the protocol has not completed.

@@

\noindent 
{\bf Notations and formalization.}  The set
of system flows is a collection ${\vec{F}}$ of
LPNs.  A {\em flow execution scenario} is defined as a
set $\{(F_{i,j}, s_{i,j})\}$ where $F_{i,j}$ is the
$j$th instance of flow $F_i \in {\vec{F}}$, and $s_{i,j}$ is
a state of $F_{i,j}$.  A flow execution scenario indicates
the set of protocols and the number of instances of a
particular protocol are activated and their corresponding
current states.  Since we assume full observability, we view
an {\em observed trace} $\rho = e_1e_2\ldots e_n$ as a
sequence of events.  Given an observed trace $\rho$, the
goal of trace interpretation is to construct a set of
candidate flow execution scenarios whose execution can
create the sequence of events in $\rho$.  We call such
execution scenarios {\em compliant} with $\rho$.  Let
$\mathit{accept(F_{i,j}, s_{i,j}, e)}$ be a function that
determines if event $e$ can be emitted by $F_{i,j}$ in state
$s_{i,j}$.  Formally, $\mathit{accept(F_{i,j}, s_{i,j}, e)}$
returns $(F_{i,j}, s'_{i,j})$ where $s'_{i,j} = (s_{i,j} - \bullet t)
\cup t\bullet$ if there exists a transition
$t$ in $F_i$ such that $L(t) = e$ and $\bullet t \subseteq
s_{i,j}$.  It returns $\emptyset$ otherwise.

%\medskip

Given an observed trace $\rho$ and the set $\vec{F}$ of LPNs,
Algorithm~\ref{algo:compliance} provides a basic procedure for
computing a set of compliant flow execution scenarios. 
The algorithm operates by keeping track (in variable
$\mathit{Scen}$) of a set of candidate flow execution scenarios
compliant with each prefix of $\rho$.  At each iteration,
for each event $e_h$ in the observed trace, we
update $Scen$ by either extending a member of
$\mathit{scen}$ or initiating a new protocol instance for 
each $scen \in Scen$ with respect to $e_h$ in every possible way.
If $e_h$ cannot be emitted by any existing or new flow instances, 
then we report that the trace is {\tt  inconsistent}, \ie, 
there is no possible interleaving of
the protocol instances from $\vec{F}$ that is compliant with
$\rho$.

\begin{algorithm}[ht]
\DontPrintSemicolon
Create an empty scenario $scen$\;
$\mathit{Scen} = \{scen\}$\;
\ForEach {$h, \; 1 \leq h \leq n $} {
	$\mathit{found} \gets {\tt true}$ \;
	$Scen' = \emptyset$\;
	\ForEach {$scen \in Scen$} {
  		\ForEach {$(F_{i,j}, s_{i,j}) \in \mathit{scen_1}$} {
    		\If {$\mathit{accept}(F_{i,j}, s_{i,j}, e_h) = (F_{i,j}, s'_{i,j})$} {
				Let $scen'$ be a copy of $scen$\;
      		$\mathit{scen'} \gets (\mathit{scen'} - (F_{i,j}, s_{i,j})) \cup (F_{i,j}, s'_{i,j})$\;
      		$\mathit{Scen'} \gets \mathit{scen'} \cup \mathit{Scen'}$\;
      		$\mathit{found} \gets {\tt false}$ \;
    		}
  		}
  		\ForEach {$F_i \in \vec{F}$} {
      	create a new instance $F_{i, j+1}$ \;
      	\If {$\mathit{accept}(F_{i,j+1},\mathit{init}_{i,j+1}, e_h) = (F_{i,j+1}, s'_{i,j+1})$} {
    			Let $scen'$ be a copy of $scen$\;
				$\mathit{scen'} \gets \mathit{scen'} \cup (F_{i,j+1}, s'_{i,j+1})$ \;
				$\mathit{Scen'} \gets \mathit{scen'} \cup \mathit{Scen'}$\;
        		$\mathit{found} \gets {\tt false}$ \;
      	}
    	}
	}
  \If {$\mathit{found} == {\tt true}$} {
    \Return ${\tt Inconsistent}$\;
  }
  $Scen = Scen'$\;
}
\Return $\mathit{Scen}$ \;
\caption{$\textsc{Check-Compliance}(\vec{F}, \, \rho)$}
\label{algo:compliance}
\end{algorithm}

To illustrate the basic idea, consider the system flow in
Fig.~\ref{flow-spec-ex}(b), which we will call $F_1$.
Suppose that the following flow trace is abstracted from an
observed signal trace.
\[
	t_1\;t_2\;t_1\;t_2\;t_3\;t_3\;t_4\;t_5\;t_5\;t_4\ldots
\]  
Here transition names in the LPN are used to represent the 
flow events in the trace.  The first four events results in
the following flow execution scenario
\[
	\{(F_{1,1}, \{p_3\}),~(F_{1,2}, \{p_3\})\}.
\]
For the first event $t_3$, it results in two execution scenarios 
below depending on which flow instance emits $t_3$.
\[
\begin{array}{l}
	\{(F_{1,1}, \{p_4\}),~(F_{1,2}, \{p_3\})\} \\
	\{(F_{1,1}, \{p_3\}),~(F_{1,2}, \{p_4\})\}.
\end{array}
\]
After handing the next event $t_3$, the above two execution scenarios
are reduced to the one as shown below.
\[
	\{(F_{1,1}, \{p_4\}),~(F_{1,2}, \{p_4\})\}.
\]
Using Algorithm~\ref{algo:compliance} to handle the
remaining four events, the following execution scenario is
derived.
\[
	\{(F_{1,1}, \{p_5, p_6\}),~(F_{1,2}, \{p_5, p_6\})\}
\]
%% Given a trace of flow events $\rho = e_1e_2\ldots e_n$, the
%% trace interpretation algorithm starts with an empty set of
%% of flow execution scenario $\mathit{Scen} = \emptyset$.
%% Then, for each $e_h$ where $1 \leq h \leq n$ starting $h=1$,
%% and for each $\mathit{scen} \in \mathit{Scen}$, the
%% following two steps are performed.
%% \begin{description}
%% \item[Step 1]~~For each $(F_{i,j}, s_{i,j}) \in
%%   \mathit{scen}$, if $\mathit{accept}(F_{i,j}, s_{i,j}, e_h)
%%   = (F_{i,j}, s'_{i,j})$, create a new scenario
%%   $\mathit{scen}' = (\mathit{scen} - (F_{i,j}, s_{i,j}))
%%   \cup (F_{i,j}, s'_{i,j})$, which is added into
%%   $\mathit{Scen}'$.

@@

In general, a signal trace of partial observability
corresponds a set of traces of flow events due to the
ambiguous interpretation of signal events.  In the
following, we discuss two cases for trace abstraction on
partial observability: mapping a single signal event to a
flow event or mapping a sequence of signal events to a flow
event.  A signal event is defined as a state on or an
assignment to a set of signals.

@@

where $\textvisiblespace$ denotes signal events that are not
mapped to any flow events.  Note that the above abstraction
leads to three distinct flow traces as the middle three
correspond to the same flow trace.

It is clear from above that a partial trace is viewed as a
set of flow traces, and Algorithm~\ref{algo:compliance} can
be suitably extended to work with flow traces to obtain the
set of candidate flows.  However, applicability of the
algorithm in practice can be gated because the number of
potential flow execution scenarios generated under partial
observability may be enormous.  Note that this is not a
limitation of the algorithm; if the observability of
critical events is poor there simply {\em are} too many flow
execution scenarios compliant with the observed trace.
Nevertheless, we need to address the issue to make trace
interpretation (whether automatic or not) practicable.
There are two potential approaches: (1)~better selection of
post-silicon trace observability, and (2)~use of system
insights during validation.  Trace signal selection itself
is an important and orthogonal topic~\cite{nicolici,basu},
and a detailed discussion of it is out of scope of this
paper.  However, we briefly describe how the debuggers'
insights of a system's architecture can help to address the
complexity issue in the trace interpretation.

@@

Post-silicon validation is performed by debuggers with deep
knowledge about the system's architecture and
microarchitecture, and the test environment.  Two key
insights are (1) the maximal number of instances of a flow
activated in the test environment, and (2)~the mutual
relationship between two flows.  For example, the test
environment may not permit multiple instances of firmware
authentication to operate concurrently, or a flow involving
audio and Web browsing to initiate until the flows
participating in boot are completed.  Our framework permits
incorporating such insights as constraints in trace
analysis; flow execution scenarios that violate these
constraints are ignored.  \textcolor{red}{These insights can lead to two advantages.
First, they help to reduce the potentially large number of partial
scenarios generated during the trace interpretation step, thus making 
the analysis more efficient.  Second, they permit the debugger
to quickly filter out uninteresting combinations of flows and focus on interesting
interleavings.}

@@

\begin{figure} 
\centerline{
\includegraphics[width=3in]{figures/Fig4.png}}
\caption{SoC platform structure.}
\label{SoC}
\end{figure}

In order to find out the efficiency of the trace analysis
method for more realistic examples, in the case study
described in this section, a more detailed transaction level
model of a SoC is constructed within the GEM5
environment~\cite{Binkert2011}.  This SoC model, as shown in Fig.~\ref{SoC},
consists of
two ARM Cortex-A9 cores, each of which contains two separate
16KB data and instruction caches.  The caches are connected
to a 1GB memory through a memory bus model.

@@

For this model, we consider the flow specifications
describing the cache coherence protocols supported in GEM5\footnote{The GEM5 cache
  coherence protocols can be found at
  \url{http://www.gem5.org/docs/html/gem5MemorySystem.html}.}
that is used to build the model in Fig~\ref{SoC}.  These
flow specifications describe data/instruction read
operations and data write operations initiated from CPUs.
Three such flows describe the cache coherent protocols for
each CPU.  As there are two CPUs in this model, there are
actually six such flows considered for this model.

@@

We wrote two simple concurrent programs, one for each CPU,
 to exercise the flows.  They read numbers from a file,
perform some operations on these numbers, and store the
results back to the file.  How GEM5 supports shared memory
multi-threaded program execution is unclear.  Therefore,
no data are shared in both caches in this test.
Furthermore, GEM5 does not support true concurrency.  When
there are two programs running on the CPUs, GEM5 alternates
the executions between the two CPUs.  To simulate
asynchronous concurrency with the interleaving semantics,
those two simple programs are instrumented with
pseudo-blocking commands, one placed before each statement.
A pseudo blocking command includes a random number generator
that returns either $0$ or $1$ and a loop that only exits
when the returned random number is $0$.

@@

The time taken to remove the irrelevant messages from the
trace is negligible.  The total runtime and the peak memory
taken by the trace analysis algorithm on the reduced
trace are 3 seconds and 12MB, respectively.  Only one flow execution
scenario is extracted, and 
Table~\ref{table-case-2} shows the number of flow instances contained in that
scenario for the six 
flows describing cache coherent operations initiated from both CPUs.
\begin{table}[tb]
\caption{The number of flow instances derived by the trace analysis with the full observability.}
\begin{center}
\begin{tabular}{|l|c|}
\hline
Flows & $\#$Instances \\
\hline
\hline
CPU1 Data Read			&  $17582$\\
CPU1 Instruction Read		&  $4002$\\
CPU1 Write				&  $3370$\\
\hline
CPU2 Data Read			&  $17386$\\
CPU2 Instruction Read		&  $3955$\\
CPU2 Write				&  $3308$\\
\hline
\end{tabular}
\end{center}
\label{table-case-2}
\end{table}%

In the second experiment, partial observability is taken into account
with the four monitors attached to the links between two CPUs and their
caches are disabled. Then, the trace is generated by the
remaining five monitors from the SoC model executing the
same program.  The new trace contains 15089 messages.  
Similarly, only one flow execution
scenario is extracted, and the numbers of the
flow instances contained in that execution scenario are
shown in Table~\ref{table-par-obs}.  From these results, the
numbers of the flow instances are dropped significantly
compared to the results extracted from the trace with the
full observability as shown in
Table~\ref{table-case-2}. This difference is due to that
some communications occurred in the system when executing
the program involve the CPUs and their corresponding caches
only, and the traffic on the links between the CPUs and
their corresponding caches is not observable. Therefore, the
instances of the flow specifications characterizing these
communications do not exist in the trace. In other words,
all extracted flow instances in Table~\ref{table-par-obs}
characterize the communications that pass through the memory
bus in the system model.  The runtime and memory usage as shown in
the third column in Table~\ref{table-results} are
similar to those for analyzing the trace of the full
observability.

@@

In the third experiment, further partial observability is taken into consideration.  In this experiment, only the five links involving the memory bus are still considered.  However, an assumption is made that all events passing the same link are not distinguishable due to the limited observability.  The monitors are modified such that whenever an event is captured on one of the links, it dumps a set of events passing through the same link into the trace file.  Therefore, each line of the trace file corresponds to a set of events.  After applying the trace analysis to this trace,  a total of 13944 flow execution scenarios are extracted.    This large number, compared to the results from the first two experiments, is due to the ambiguous interpretation of the events with limited observability.

The whole experiment takes about $15$ minutes and $420$~MB to finish as shown in column~4
in Table~\ref{table-results}, significantly higher than the numbers for analyzing traces where there is no ambiguity in the observed events.  This is due to the fact that a trace of ambiguous events is in fact a set of traces of original events, which lead to large numbers of execution scenarios either during or at the end of the analysis.  In this experiment, the peak number of execution scenarios during the analysis process is $70384$, many of which are invalid and removed eventually.  However, controlling the number of intermediate execution scenarios during the trace analysis is critical in order for the analysis to be tractable.  Here, insights from validators could help, but are not used in this experiment.
